---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 22 @ 11:59PM
author: "Puyuan Liu, 806152535"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. 
You should use the `los_long` variable as the outcome. 
Your algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms. 

1. Data preprocessing and feature engineering.

```{r}
library(tidyverse)
library(readr)
library(gtsummary)
library(tidymodels)
library(glmnet)
library(ranger)
library(xgboost)
library(stacks)
library(vip)
```

```{r}
icu_cohort <- read_rds("~/203b-hw/hw4/mimiciv_shiny/mimic_icu_cohort.rds")
```

```{r}
# Define a function to removes out-liers based on IQR
remove_outliers_iqr <- function(data_vector) {
  # Calculate the first and third quartiles (Q1, Q3) and the IQR
  quartiles <- quantile(data_vector, probs = c(0.25, 0.75), na.rm = TRUE)
  iqr_value <- IQR(data_vector, na.rm = TRUE)
  # Replace out-liers with NA
  cleaned_vector <- dplyr::if_else(
    data_vector < quartiles[1] - 1.5 * iqr_value | 
    data_vector > quartiles[2] + 1.5 * iqr_value, 
    NA, 
    data_vector
  )
  # Return the vector with outliers replaced by NA
  return(cleaned_vector)
}


# Data preprocessing and feature engineering
mimic_icu_cohort <- icu_cohort |>
  
  # Select columns of interest
  select(subject_id, hadm_id, stay_id, los_long, first_careunit, admission_type,
         marital_status, race, gender, age, insurance, 
         Sodium, Chloride, Creatinine, Potassium, Glucose, Hematocrit, 
         `White Blood Cells`, Bicarbonate,
         `Temperature Fahrenheit`, `Non Invasive Blood Pressure diastolic`, 
         `Respiratory Rate`, `Non Invasive Blood Pressure systolic`, 
         `Heart Rate`)|>
  
  # Convert character and logical columns to factors
  mutate_if(is.character, as.factor) |>
  mutate_if(is.logical, as.factor) |>
  
  # Apply the remove_outliers_iqr function to numeric columns
  mutate_if(is.numeric, remove_outliers_iqr) |>
  
  # Impute NA's in numerical columns with the median
  mutate(across(where(is.numeric), 
                ~if_else(is.na(.), 
                         median(., na.rm = TRUE), .))) |>
  print(width = Inf)
```


2. Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.
```{r}
set.seed(203)

# sort
mimiciv_icu_cohort <- mimic_icu_cohort |>
  arrange(subject_id, hadm_id, stay_id)|>
  select(-c(subject_id, hadm_id, stay_id))

data_split <- initial_split(
  mimiciv_icu_cohort, 
  # stratify by los_long
  strata = "los_long", 
  prop = 0.5
  )
```

```{r}
summary(mimiciv_icu_cohort)
```

```{r}
icu_train <- training(data_split)
dim(icu_train)
```
```{r}
icu_test <- testing(data_split)
dim(icu_test)
```

```{r}
data_split |>
  write_rds("data_split.rds", compress = "gz")
```


3. Train and tune the models using the training set.

### Logistic regression with enet regularization

```{r}
logit_recipe <- 
  recipe(
    los_long ~ ., 
    data = icu_train
  ) |>
  # mode imputation for categorical_var
  step_impute_mode(all_nominal_predictors()) |>
  # create traditional dummy variables
  step_dummy(all_nominal_predictors()) |>
  # zero-variance filter
  step_zv(all_numeric_predictors()) |> 
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) 
logit_recipe
```


```{r}
# Models
logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) |> 
  set_engine("glmnet", standardize = FALSE) 
logit_mod
```

```{r}
# Workflow
logit_wf <- workflow() |>
  add_recipe(logit_recipe) |>
  add_model(logit_mod) 
logit_wf
```

```{r}
# Tuning grid
logit_param_grid <- grid_regular(
  penalty(range = c(-2.5, -2)), 
  mixture(),
  levels = c(5, 2)
  ) 
logit_param_grid
```

```{r}
set.seed(203)

folds <- vfold_cv(icu_train, v = 3)
folds
```

```{r}
# Fit cross-validation (logistic)
if (file.exists("logit_fit.rds")) {
  logit_fit <- read_rds("logit_fit.rds")
} else {
  logit_fit <- logit_wf |>
    tune_grid(
      resamples = folds,
      grid = logit_param_grid,
      metrics = metric_set(roc_auc, accuracy),
      control = control_stack_grid()
      ) 
  write_rds(logit_fit, "logit_fit.rds")
}
logit_fit
```

```{r}
# Visualize CV results
logit_fit |>
  # aggregate metrics from K folds
  collect_metrics() |>
  #print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = penalty, y = mean, color = factor(mixture))) +
  geom_point() +
  labs(x = "Penalty", y = "CV AUC") +
  scale_x_log10()
```
```{r}
logit_fit |>
  show_best("roc_auc")
```

```{r}
best_logit <- logit_fit |>
  select_best("roc_auc")
best_logit
```

```{r}
# Final workflow
logit_final_wf <- logit_wf |>
  finalize_workflow(best_logit)
logit_final_wf
```

```{r}
# Fit the whole training set, then predict the test cases
if (file.exists("logit_final_fit.rds")) {
  logit_final_fit <- read_rds("logit_final_fit.rds")
} else {
  logit_final_fit <- logit_final_wf |>
    last_fit(data_split) |>
    write_rds("logit_final_fit.rds")
}
logit_final_fit
```

```{r}
# Test metrics
logit_final_fit |> 
  collect_metrics()
```


### Random forest

```{r}
rf_recipe <- 
  recipe(
    los_long ~ ., 
    data = icu_train
  ) |>
  # mode imputation for categorical_var
  step_impute_mode(all_nominal_predictors()) |>
  # zero-variance filter
  step_zv(all_numeric_predictors()) 
rf_recipe
```

```{r}
# Model
rf_mod <- 
  rand_forest(
    mode = "classification",
    # Number of predictors randomly sampled in each split
    mtry = tune(),
    # Number of trees in ensemble
    trees = tune()
  ) |> 
  set_engine("ranger", importance = "impurity") 
rf_mod
```

```{r}
# Workflow
rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_mod) 
rf_wf
```

```{r}
# Tuning grid
rf_param_grid <- grid_regular(
  trees(range = c(400L, 600L)), 
  mtry(range = c(1L, 3L)),
  levels = c(3, 3)
  )
rf_param_grid
```

```{r}
# Fit cross-validation (boosting)
set.seed(203)
if (file.exists("rf_fit.rds")) {
  rf_fit <- read_rds("rf_fit.rds")
} else {
  rf_fit <- rf_wf |>
    tune_grid(
      resamples = folds,
      grid = rf_param_grid,
      metrics = metric_set(roc_auc, accuracy),
      control = control_stack_grid()
      )
  write_rds(rf_fit, "rf_fit.rds")
}
rf_fit
```

```{r}
# Visualize CV results
rf_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = trees, y = mean, color = factor(mtry))) +
  geom_point() + 
  labs(x = "Num. of Trees", y = "CV AUC")
```

```{r}
# Show the best 5 model
rf_fit |>
  show_best("roc_auc")
```

```{r}
# Select the best model
best_rf <- rf_fit |>
  select_best("roc_auc")
best_rf
```

```{r}
# Final workflow
rf_final_wf <- rf_wf |>
  finalize_workflow(best_rf)
rf_final_wf
```

```{r}
# Fit the whole training set, then predict the test cases
if (file.exists("rf_final_fit.rds")) {
  rf_final_fit <- read_rds("rf_final_fit.rds")
} else {
  rf_final_fit <- rf_final_wf |>
    last_fit(data_split)
  write_rds(rf_final_fit, "rf_final_fit.rds")
}

rf_final_fit
```

```{r}
# Test metrics
rf_final_fit |> 
  collect_metrics()
```

### Boosting

```{r}
boost_recipe <- 
  recipe(
    los_long ~ ., 
    data = icu_train
  ) |>
  # mode imputation for categorical_var
  step_impute_mode(all_nominal_predictors()) |>
  # create traditional dummy variables (necessary for xgboost)
  step_dummy(all_nominal_predictors()) |>
  # zero-variance filter
  step_zv(all_numeric_predictors()) 
boost_recipe
```

```{r}
# Model
boost_mod <- 
  boost_tree(
    mode = "classification",
    trees = 400,
    tree_depth = tune(),
    learn_rate = tune()
  ) |> 
  set_engine("xgboost") 
boost_mod
```

```{r}
# Workflow
boost_wf <- workflow() |>
  add_recipe(boost_recipe) |>
  add_model(boost_mod)
boost_wf
```

```{r}
# Tuning grid
boost_param_grid <- grid_regular(
  tree_depth(range = c(2L, 3L)),
  learn_rate(range = c(-2, 0), trans = log10_trans()),
  levels = c(2, 5)
  )
boost_param_grid
```

```{r}
# Fit cross-validation (boosting)
set.seed(203)
if (file.exists("boost_fit.rds")) {
  boost_fit <- read_rds("boost_fit.rds")
} else {
  boost_fit <- boost_wf |>
    tune_grid(
      resamples = folds,
      grid = boost_param_grid,
      metrics = metric_set(roc_auc, accuracy),
      control = control_stack_grid()
      ) 
  write_rds(boost_fit, "boost_fit.rds")
}
boost_fit
```

```{r}
# Visualize CV results
boost_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = learn_rate, y = mean, color = factor(tree_depth))) +
  geom_point() + 
  labs(x = "Learning Rate", y = "CV AUC") +
  scale_x_log10()
```

```{r}
# Show the best 5 model
boost_fit |>
  show_best("roc_auc")
```

```{r}
# Select the best model
best_boost <- boost_fit |>
  select_best("roc_auc")
best_boost
```

```{r}
# Final workflow
boost_final_wf <- boost_wf |>
  finalize_workflow(best_boost)
boost_final_wf
```

```{r}
# Fit the whole training set, then predict the test cases
if (file.exists("boost_final_fit.rds")) {
  boost_final_fit <- read_rds("boost_final_fit.rds")
} else {
  boost_final_fit <- boost_final_wf |>
    last_fit(data_split) |>
    write_rds("boost_final_fit.rds")
}
boost_final_fit
```

```{r}
# Test metrics
boost_final_fit |> 
  collect_metrics()
```

### Model stacking

```{r}
stack_recipe <- 
  recipe(
    los_long ~ ., 
    data = icu_train
  ) |>
  # mode imputation for categorical_var
  step_impute_mode(all_nominal_predictors()) |>
  # create traditional dummy variables (necessary for svm)
  step_dummy(all_nominal_predictors()) |>
  # zero-variance filter
  step_zv(all_numeric_predictors()) |>
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) 
stack_recipe
```

```{r}
# Model stacking
if (!file.exists("icu_model_st.rds")) {
  icu_model_st <- 
    # initialize the stack
    stacks() |>
    # add candidate members
    add_candidates(logit_fit) |>
    add_candidates(rf_fit) |>
    add_candidates(boost_fit) |>
    # determine how to combine their predictions
    blend_predictions(
      penalty = 10^(-4:-1),
      metrics = c("roc_auc"),
      times = 3
      ) |>
    # fit the candidates with nonzero stacking coefficients
    fit_members()
  saveRDS(icu_model_st, "icu_model_st.rds")
} else {
  icu_model_st <- read_rds("icu_model_st.rds")
}
```

```{r}
icu_model_st
```

```{r}
autoplot(icu_model_st)
```

```{r}
# To see the top results
autoplot(icu_model_st, type = "weights")
```

```{r}
# To identify which model configurations were assigned what stacking coefficients
collect_parameters(icu_model_st, "rf_fit")
```
```{r}
collect_parameters(icu_model_st, "logit_fit")
```

```{r}
collect_parameters(icu_model_st, "boost_fit")
```

```{r}
# Final classfication for stacking method
stack_icu_pred <- icu_test %>%
  bind_cols(predict(icu_model_st, ., type = "prob")) %>%
  print(width = Inf)
```

4. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?

```{r}
# Test metrics for logistic regression model
logit_final_fit |> 
  collect_metrics()
```

```{r}
# Test metrics for random forest model
rf_final_fit |> 
  collect_metrics()
```

```{r}
# Test metrics for boosting model
boost_final_fit |> 
  collect_metrics()
```

```{r}
# Convert prediction estimates to factor based on the chosen threshold
stack_icu_pred <- stack_icu_pred %>%
  mutate(
    .pred_class = if_else(.pred_FALSE >= 0.5, "FALSE", "TRUE"),
    .pred_class = factor(.pred_class, levels = c("FALSE", "TRUE"))
  )

# Test metrics for stacking model
roc_auc_value <- yardstick::roc_auc(
  stack_icu_pred,
  truth = los_long,
  contains(".pred_FALSE")
)

accuracy_value <- yardstick::accuracy(
  stack_icu_pred,
  truth = los_long,
  estimate = .pred_class
)

# Print both metrics
print(roc_auc_value)
print(accuracy_value)
```

```{r}
# Extract important features for logistics regression model 
logit_features <- extract_fit_engine(logit_final_fit) |>
  vip(num_features = 6)
logit_features
```

```{r}
# Extract important features for random forest model
rf_features <- extract_fit_engine(rf_final_fit) |>
  vip(num_features = 6)
rf_features
```

```{r}
# Extract important features for boosting model
boost_features <- extract_fit_engine(boost_final_fit) |>
  vip(num_features = 6)
boost_features
```


